High-order cognition is supported by information-rich but compressible brain activity patterns

### 2023-03-16 paper/main.tex
finished complete drafts of main text and supplement

To the extent that brain activity patterns contain rich task-relevant
information, we should be able to use the activity patterns to accurately
differentiate between different aspects of the task~\citep[e.g., using pattern
classifiers;][]{NormEtal06b}. For example, prior work has shown a direct
correspondence between classification accuracy and the information content of a
signal~\citep{Alva02}. To the extent that brain activity patterns are
compressible, we should be able to generate simplified (e.g., lower
dimensional) representations of the data while still preserving the relevant or
important aspects of the original signal. In general, information content and
compressibility are related but are partially dissociable
(Fig.~\ref{fig:information-compression}). If a given signal (e.g., a
representation of brain activity patterns) contains more information about
ongoing cognitive processes, then the peak decoding accuracy should be high.
And if the signal is compressible, a low-dimensional embedding of the signal
will be similarly informative to the original signal
(Fig.~\ref{fig:information-compression}D).

@@

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. For each condition, decoding accuracy has been normalized
by subtracting expected ``chance'' decoding performance (estimated as
$\frac{1}{T}$, where $T$ is the number of timepoints in the given condition).
This normalization was intended to enable fairer comparisons across conditions;
a relative decoding accuracy of 0.0 denotes ``chance'' performance. The
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows. All
error ribbons in Panels A and B denote bootstrap-estimated 95\% confidence
intervals. \textbf{C. Estimating ``compressibility'' for each condition.} The
density plots display the numbers of components required to reach at least 5\%
(corrected) decoding accuracy, across all cross validation runs. For the
``rest'' condition, where decoding accuracy never reached 5\% accuracy, we
display the distribution of the numbers of components required to achieve the
peak decoding accuracy in each fold. (This distribution is nearly flat, as
illustrated in Panel E.) \textbf{D. Estimating ``informativeness'' for each
condition.} The density plots display the peak (corrected) decoding accuracies
achieved in each condition, across all cross validation runs. \textbf{E.
Informativeness versus compressibility.} Each dot displays the minimum number
of components required to achieve at least 5\% corrected decoding accuracy
($x$-coordinate; for the rest condition the numbers of components are estimated
using the peak decoding accuracies as in Panel C) and the peak (corrected)
decoding accuracy ($y$-coordinate). The smaller dots correspond to individual
cross validation runs and the larger dots display the averages across runs. The
inset displays a zoomed-in view of the indicated region in the main panel.}

@@

The cognitively richer conditions also displayed steeper initial slopes. For
example, the intact condition decoders reached an arbitrarily chosen threshold
of 5\% accuracy using fewer components than the paragraph condition decoders
($t(99) = -7.429, p < 0.001)$) or word condition decoders ($t(99) = -7.300, p <
0.001)$), and decoding accuracy never exceeded 5\% for the rest condition. This
suggests that brain activity patterns evoked by cognitively richer conditions
are more compressible, such that representing the data using the same number of
principal components provides more information to the temporal decoders
(Figs.~\ref{fig:inflection}B, C). Taken together, as shown in
Figure~\ref{fig:inflection}E, we found that brain activity patterns evoked by
cognitively richer conditions tended to be both more informative (i.e.,
associated with higher decoding accuracies) \textit{and} more compressible
(i.e., requiring fewer components to achieve the 5\% accuracy threshold).

@@

If informativeness (to the temporal decoders) and compressibility vary with the
cognitive richness of the stimulus, might these measures also vary over time
\textit{within} a given condition? For example, participants in the intact
condition might process the ongoing story more deeply later on in the story
(compared with earlier in the story) given the additional narrative background
and context they had been exposed to by that point. To examine this
possibility, we divided each condition into four successive time segments. We
computed decoding curves (Fig.~\ref{fig:inflection-quarters}A) and the numbers
of components required to achieve 5\% decoding accuracy
(Fig.~\ref{fig:inflection-quarters}B) for each segment and condition. We found
that, in the two most cognitively rich conditions (intact and paragraph), both
decoding accuracy and compressibility, as reflected by the change in decoding
curves, increased with listening time (e.g., at the annotated reference point
of $k = 20$ components in Fig.~\ref{fig:inflection-quarters}C: intact: $t(99) =
7.915, p < 0.001)$; paragraph: $t(99) = 2.354, p = 0.021)$). These changes may
reflect an increase in comprehension or depth of processing with listening
time. In contrast, the decoding accuracy and compressibility \textit{decreased}
with listening time in the word condition ($t(99) = -10.747, p < 0.001)$) and
rest condition ($t(99) = -22.081, p < 0.001)$). This might reflect the
depletion of attentional resources in the less-engaging word and rest
conditions.

@@

To gain further insights into which brain functions might be most closely
associated with the top-weighted components from each experimental condition,
we manually grouped each Neurosynth-derived topic into a smaller set of core
cognitive functions. Separately for each component, we computed the average
weightings across all topics that were tagged as being associated with each of
these cognitive functions (Figs.~\ref{fig:neurosynth-summary}A,~\neurosynthFull
A). To help visualize these associations, we used the patterns of associations
for each component to construct graphs whose nodes were experimental conditions
and cognitive functions (Figs.~\ref{fig:neurosynth-summary}B,~\neurosynthFull
B). We also computed correlations between the sets of per-topic weightings from
each of the top-weighted components from each experimental condition
(Fig.~\ref{fig:neurosynth-summary}C, D) and between the brain maps for each
condition's components (Fig.~\neurosynthFull C, D). Taken together, we found
that each component appeared to weight on a fundamental set of cognitive
functions that varied by experimental condition. For example, the top principal
components from every condition weighted similarly on the full set of
Neurosynth topics (Fig.~\ref{fig:neurosynth-pca}A) and cognitive functions
(Figs.~\ref{fig:neurosynth-summary}A, B and~\neurosynthFull~A, B), suggesting
that these components might reflect a set of functions or activity patterns
that are common across all conditions. The second components' weightings were
similar across the intact, paragraph, and rest conditions (highest-weighted
functions: cognitive control, memory, social cognition, and resting state), but
different for the word condition (highest-weighted functions: sensory
perception and cognitive control). The fourth components' weighting were
grouped the paragraph and word conditions (highest-weighted functions: memory,
language processing, and cognitive control) and the intact and rest conditions
(highest-weighted functions: emotion, social cognition). We also used
ChatGPT~\citep{ChatGPT} to sort the list of manually tagged cognitive functions
from lowest-level to highest-level (Tab.~\topicTags,
Fig.~\ref{fig:neurosynth-summary}E; also see \textit{Ranking cognititive
processes}). We found that higher-level functions tended to be weighted on more
heavily by top components from the intact and paragraph conditions than
lower-level functions. The top components from the word condition showed the
opposite tendancy, whereby \textit{lower}-level functions tended to be weighted
on more heavily than higher-level functions. The components from the rest
condition showed almost no differences in the weights associated with
high-level versus low-level functions.

@@

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and several core cognitive function classifications is
displayed in the lower right of the figure. Table~\topics~provides a list of
each topic's top-weighted terms, along with each topic's manually labeled
cognitive classification. A full list of the topics most highly associated with
each component may be found in Figure~\topTerms. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heatmaps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heatmaps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition components, by
component.} Each sub-panels network plot summarizes the pattern of correlations
between the $n$\textsuperscript{th} top-weighted principal components
(sub-panel) for each experimental condition (gray-outlined dots). The line
thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heatmaps in Panel C). \textbf{E. Change in
weights by condition.} The bar heights reflect the slopes of regression lines,
fit separately to the top five components from each condition, between the
ChatGPT-derived ``rank'' of each cognitive classification and the correlations
between the component and topic maps associated with cognitive processes at the
given rank (see \textit{Ranking cognitive processes}).  Also see
Fig.~\neurosynthFull~for additional information.}

@@

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns evoked by cognitively richer stimuli were also more
compressible, in that each individual component provided more ``signal'' to
temporal decoders relative to components of data from less cognitively rich
conditions (Fig.~\ref{fig:inflection}). Over time (e.g., as the experiment
progressed), these phenomena were strengthened. Specifically, across story
segments, data from more cognitively rich conditions became more informative
and compressible, and data from less cognitively rich conditions became
\textit{less} informative and compressible (Fig.~\ref{fig:inflection-quarters}).
We also repeated these analyses separately for different brain networks. We
found that networks traditionally associated with higher-level cognitive
functions tended to provide more informative brain patterns than networks
traditionally associated with lower level cognitive functions
(Fig.~\ref{fig:networks}). Finally, we examined the most dominant components of
the brain activity patterns from each experimental condition. We used a reverse
inference approach~\citep{RubiEtal17} to identify the terms in the neuroimaging
literature most commonly associated with the corresponding maps. As summarized
in Figure~\ref{fig:neurosynth-summary}, we found that terms associated with memory and
sensory processing were associated with the strongest components in all three
story listening conditions. Terms associated with sensory integration were
associated with the strongest components in the intact and paragraph-scrambled
conditions. Terms associated with sentence comprehension, emotion, and valence
were associated with the strongest components in the intact condition. Finally,
terms associated with the default mode network were associated with the
strongest components in the word-scrambled and resting state conditions. Taken
together, our findings indicate that the informativeness and compressibility of
our brain activity patterns are task-dependent, and these properties change
systematically with factors like cognitive richness and depth of processing.

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10, SizeEtal18,
SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11, AdacEtal12,
BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04, DhamEtal08,
KorzEtal08, BrovEtal04}. Correlations or causal associations between different
brain regions simultaneously imply that full-brain activity patterns will be
compressible and also that those activity patterns will contain redundancies.
For example, the extent ot which activity patterns at one brain area can be
inferred or predicted from activity patterns at other
areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then the
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region.

@@

Following our prior related work, we used HTFA~\citep{MannEtal18} to derive a
compact representation of the neuroimaging data. In brief, this approach
approximates the timeseries of voxel activations (44,415 voxels) using a much
smaller number of radial basis function (RBF) nodes~\citep[in this case, 700
nodes, as determined by an optimization procedure;][]{MannEtal18}). This
provides a convenient representation for examining full-brain activity patterns
and network dynamics. All of the analyses we carried out on the neuroimaging
dataset were performed in this lower-dimensional space. In other words, each
participant's data matrix was a number-of-timepoints ($T$) by 700 matrix of
HTFA-derived factor weights (where the row and column labels were matched
across participants). Code for carrying out HTFA on fMRI data may be found as
part of the BrainIAK toolbox~\citep{CapoEtal17, KumaEtal21}, which may be
downloaded at \href{https://brainiak.org/}{brainiak.org}.

@@

To help interpret the brain activity patterns we found within the contexts of
other studies, we created summary maps of each principal component, for each
experimental condition. Each principal component comprises 700 ``weights'' on
each of the HTFA-derived RBF nodes (see \textit{Hierarchical Topographic Factor
Analysis}). For each node, we evaluated its RBF at the locations of every voxel
in the standard 2~mm MNI152 template brain and multiplied the RBF by the node's
weight. The sum of these weighted RBF activation maps provides a full-brain
image, in MNI152 space, of the given principal component.

Next, we considered 80 topics estimated using Latent Dirichlet
Allocation~\citep{BleiEtal03} applied to 9,204 functional neuroimaging articles
in the Neurosynth database~\citep{RubiEtal17}. The topics, as well as
associated brain maps identified using Neurosynth, were identified and reported
in several prior studies~\citep{FoxEtal14, SulEtal17, ChenEtal20}. The topic
labels for each topic were generated automatically with the following ChatGPT
(chat.openai.com) prompt: ``Please help me come up with intuitive labels for
topic topics I found by fitting a topic model to thousands of neuroscience and
psychology articles. I'll paste in the top 10 highest-weighted words for each
topic, and I'd like you to respond with a suggested label. For each topic,
please respond with just the topic label and no other formatting or text. Here
are the next topic's top words:'' followed by a comma-separated list of the
given topic's top-weighted words reflected in the Table~\topics. For some
topics, ChatGPT responded with a longer-form response rather than a concise
topic label. In these instances, on a case-by-case basis, we used a second
follow-up prompt to achieve the given topic's label: ``could you please come up
with a more concise label for that topic?''. We then manually identified a set
of 11 cognitive labels that were intended to encapsulate a representative range
of widely studied low-level and high-level cognitive functions. In choosing the
set of cognitive labels, we jointly considered each topic's ChatGPT-derived
topic label, along with the top-weighted words for the topic. We attempted to
generate a concise set of labels that still spanned the full set of cognitive
functions reflected across the 80 topics. Topics that appeared unrelated to
specific cognitive functions (e.g., topics related to specific methods or
clinical themes) are designated with dashes in Table~\topics.

@@

We manually identified 11 cognitive labels
spanning the set of 80 Neurosynth-derived topics: cognitive control, language
processing, memory, emotion, social cognition, spatial cognition, attention,
reward, sensory perception, motor control, and resting state. We then used
ChatGPT to automatically ``rank'' the processes from high-level to low-level
using the following prompt: ``Please rank these cognitive processes from
highest-level to lowest-level, where higher values indicate higher-order or
higher-level processes. return the result as a csv file with a header row and
two columns: `Cognitive label' and `Rank'. Here are the processes: cognitive
control, language processing, memory, emotion, social cognition, spatial
cognition, attention, reward, sensory perception, motor control, resting
state''.  Table~\topicTags displays the output.  We used these labels in
the analysis presented in Figure~\ref{fig:neurosynth-summary}E to help summarize
difference in topic weightings across experimental conditions.

>>>>

To the extent that brain activity patterns contain rich task-relevant
information, we should be able to use the activity patterns to accurately
differentiate between different aspects of a task~\citep[e.g., using pattern
classifiers;][]{NormEtal06b}. For example, prior work has shown a direct
correspondence between classification accuracy and the information content of a
signal~\citep{Alva02}. To the extent that brain activity patterns are
compressible, we should be able to generate simplified (e.g., lower
dimensional) representations of the data while still preserving the relevant or
important aspects of the original signal. In general, information content and
compressibility are related but are partially dissociable
(Fig.~\ref{fig:information-compression}). If a given signal (e.g., a
representation of brain activity patterns) contains more information about
ongoing cognitive processes, then the peak decoding accuracy should be high.
And if the signal is compressible, a low-dimensional embedding of the signal
will be similarly informative as the original signal
(Fig.~\ref{fig:information-compression}D).

@@

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. For each condition, decoding accuracy has been normalized
by subtracting expected ``chance'' decoding performance (estimated as
$\frac{1}{T}$, where $T$ is the number of timepoints in the given condition).
This normalization was intended to enable fairer comparisons across conditions;
a relative decoding accuracy of 0.0 denotes ``chance'' performance. The
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows. All
error ribbons in Panels A and B denote bootstrap-estimated 95\% confidence
intervals. \textbf{C. Estimating ``compressibility'' for each condition.} The
density plots display the numbers of components required to reach at least 5\%
(corrected) decoding accuracy, across all cross validation runs. For the
``rest'' condition, where decoding accuracy never reached 5\% accuracy, we
display the distribution of the numbers of components required to achieve the
peak decoding accuracy in each fold. (This distribution is nearly flat, as
also illustrated in Panel E.) \textbf{D. Estimating ``informativeness'' for each
condition.} The density plots display the peak (corrected) decoding accuracies
achieved in each condition, across all cross validation runs. \textbf{E.
Informativeness versus compressibility.} Each dot displays the minimum number
of components required to achieve at least 5\% corrected decoding accuracy
($x$-coordinate; for the rest condition the numbers of components are estimated
using the peak decoding accuracies as in Panel C) and the peak (corrected)
decoding accuracy ($y$-coordinate). The smaller dots correspond to individual
cross validation runs and the larger dots display the averages across runs. The
inset displays a zoomed-in view of the indicated region in the main panel.}

@@

The cognitively richer conditions also displayed steeper initial slopes. For
example, the intact condition decoders reached an arbitrarily chosen threshold
of 5\% accuracy using fewer components than the paragraph condition decoders
($t(99) = -7.429, p < 0.001$) or word condition decoders ($t(99) = -7.300, p <
0.001$), and decoding accuracy never exceeded 5\% for the rest condition. This
suggests that brain activity patterns evoked by cognitively richer conditions
are more compressible, such that representing the data using the same number of
principal components provides more information to the temporal decoders
(Figs.~\ref{fig:inflection}B, C). Taken together, as shown in
Figure~\ref{fig:inflection}E, we found that brain activity patterns evoked by
cognitively richer conditions tended to be both more informative (i.e.,
associated with higher decoding accuracies) \textit{and} more compressible
(i.e., requiring fewer components to achieve the 5\% accuracy threshold).

@@

If informativeness (to the temporal decoders) and compressibility vary with the
cognitive richness of the stimulus, might these measures also vary over time
\textit{within} a given condition? For example, participants in the intact
condition might process the ongoing story more deeply later on in the story
(compared with earlier in the story) given the additional narrative background
and context they had been exposed to by that point. To examine this
possibility, we divided each condition into four successive time segments. We
computed decoding curves (Fig.~\ref{fig:inflection-quarters}A) and the numbers
of components required to achieve 5\% decoding accuracy
(Fig.~\ref{fig:inflection-quarters}B) for each segment and condition. We found
that, in the two most cognitively rich conditions (intact and paragraph), both
decoding accuracy and compressibility, as reflected by the change in decoding
curves, increased with listening time (e.g., at the annotated reference point
of $k = 20$ components in Fig.~\ref{fig:inflection-quarters}C: intact: $t(99) =
7.915, p < 0.001$; paragraph: $t(99) = 2.354, p = 0.021$). These changes may
reflect an increase in comprehension or depth of processing with listening
time. In contrast, the decoding accuracy and compressibility \textit{decreased}
with listening time in the word condition ($t(99) = -10.747, p < 0.001$) and
rest condition ($t(99) = -22.081, p < 0.001$). This might reflect the
depletion of attentional resources in the less-engaging word and rest
conditions.

@@

To gain further insights into which brain functions might be most closely
associated with the top-weighted components from each experimental condition,
we manually grouped each Neurosynth-derived topic into a smaller set of core
cognitive functions. Separately for each component, we computed the average
weightings across all topics that were tagged as being associated with each of
these cognitive functions (Figs.~\ref{fig:neurosynth-summary}A,~\neurosynthFull
A). To help visualize these associations, we used the patterns of associations
for each component to construct graphs whose nodes were experimental conditions
and cognitive functions (Figs.~\ref{fig:neurosynth-summary}B,~\neurosynthFull
B). We also computed correlations between the sets of per-topic weightings from
each of the top-weighted components from each experimental condition
(Fig.~\ref{fig:neurosynth-summary}C, D) and between the brain maps for each
condition's components (Fig.~\neurosynthFull C, D). Taken together, we found
that each component appeared to weight on a fundamental set of cognitive
functions that varied by experimental condition. For example, the top principal
components from every condition weighted similarly on the full set of
Neurosynth topics (Fig.~\ref{fig:neurosynth-pca}A) and cognitive functions
(Figs.~\ref{fig:neurosynth-summary}A, B and~\neurosynthFull~A, B), suggesting
that these components might reflect a set of functions or activity patterns
that are common across all conditions. The second components' weightings were
similar across the intact, paragraph, and rest conditions (highest-weighted
functions: cognitive control, memory, social cognition, and resting state), but
different for the word condition (highest-weighted functions: sensory
perception and cognitive control). The fourth components' weighting 
grouped the paragraph and word conditions (highest-weighted functions: memory,
language processing, and cognitive control) and the intact and rest conditions
(highest-weighted functions: emotion, social cognition). We also used
ChatGPT~\citep{ChatGPT} to sort the list of manually tagged cognitive functions
from lowest-level to highest-level (Tab.~\topicTags,
Fig.~\ref{fig:neurosynth-summary}E; also see \textit{Ranking cognititive
processes}). We found that higher-level functions tended to be weighted on more
heavily by top components from the intact and paragraph conditions than
lower-level functions. The top components from the word condition showed the
opposite tendancy, whereby \textit{lower}-level functions tended to be weighted
on more heavily than higher-level functions. The components from the rest
condition showed almost no differences in the weights associated with
high-level versus low-level functions.

@@

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and several core cognitive function classifications is
displayed in the lower right of the figure. Table~\topics~provides a list of
each topic's top-weighted terms, along with each topic's manually labeled
cognitive classification. A full list of the topics most highly associated with
each component may be found in Figure~\topTerms. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heatmaps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heatmaps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition components, by
component.} Each sub-panel's network plot summarizes the pattern of correlations
between the $n$\textsuperscript{th} top-weighted principal components
(sub-panel) for each experimental condition (gray-outlined dots). The line
thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heatmaps in Panel C). \textbf{E. Change in
weights by condition.} The bar heights reflect the slopes of regression lines,
fit separately to the top five components from each condition, between the
ChatGPT-derived ``rank'' of each cognitive classification and the correlations
between the component and topic maps associated with cognitive processes at the
given rank (see \textit{Ranking cognitive processes}).  Also see
Fig.~\neurosynthFull~for additional information.}

@@

We examined fMRI data collected as participants listened to an auditory
recording of a story, scrambled recordings of the story, or underwent a resting
state scan. We found that cognitively richer stimuli evoked more reliable
(i.e., consistent across people) and information rich brain activity patterns.
The brain patterns evoked by cognitively richer stimuli were also more
compressible, in that each individual component provided more ``signal'' to
temporal decoders relative to components of data from less cognitively rich
conditions (Fig.~\ref{fig:inflection}). Over time (e.g., as the experiment
progressed), these phenomena were strengthened. Specifically, across story
segments, data from more cognitively rich conditions became more informative
and compressible, and data from less cognitively rich conditions became
\textit{less} informative and compressible
(Fig.~\ref{fig:inflection-quarters}). We also repeated these analyses
separately for different brain networks. We found that networks traditionally
associated with higher-level cognitive functions tended to provide more
informative brain patterns than networks traditionally associated with lower
level cognitive functions (Fig.~\ref{fig:networks}). Finally, we examined the
most dominant components of the brain activity patterns from each experimental
condition. We used a reverse inference approach~\citep{RubiEtal17} to identify
the terms in the neuroimaging literature most commonly associated with the
corresponding maps. As summarized in Figure~\ref{fig:neurosynth-summary}E, we
found that the intact and paragraph conditions tended to weight on higher-level
cognitive processes more than lower-level cognitive processes, whereas the word
condition weighted on lower-level processes more than higher-level processes
and the rest condition showed no difference in high-level versus low-level
weighting. Taken together, our findings indicate that the informativeness and
compressibility of our brain activity patterns are task-dependent, and these
properties change systematically with factors like cognitive richness and depth
of processing.

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10, SizeEtal18,
SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11, AdacEtal12,
BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04, DhamEtal08,
KorzEtal08, BrovEtal04}. Correlations or causal associations between different
brain regions simultaneously imply that full-brain activity patterns will be
compressible and also that those activity patterns will contain redundancies.
For example, the extent to which activity patterns at one brain area can be
inferred or predicted from activity patterns at other
areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then a
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region.

@@

Following our prior related work, we used HTFA~\citep{MannEtal18} to derive a
compact representation of the neuroimaging data. In brief, this approach
approximates the timeseries of voxel activations (44,415 voxels) using a much
smaller number of radial basis function (RBF) nodes~\citep[in this case, 700
nodes, as determined by an optimization procedure;][]{MannEtal18}. This
provides a convenient representation for examining full-brain activity patterns
and network dynamics. All of the analyses we carried out on the neuroimaging
dataset were performed in this lower-dimensional space. In other words, each
participant's data matrix was a number-of-timepoints ($T$) by 700 matrix of
HTFA-derived factor weights (where the row and column labels were matched
across participants). Code for carrying out HTFA on fMRI data may be found as
part of the BrainIAK toolbox~\citep{CapoEtal17, KumaEtal21}, which may be
downloaded at \href{https://brainiak.org/}{brainiak.org}.

@@

To help interpret the brain activity patterns we found within the contexts of
other studies, we created summary maps of each principal component, for each
experimental condition. Each principal component comprises 700 ``weights'' on
each of the HTFA-derived RBF nodes (see \textit{Hierarchical Topographic Factor
Analysis}). For each node, we evaluated its RBF at the locations of every voxel
in the standard 2~mm MNI152 template brain and multiplied the RBF by the node's
weight. The sum of these weighted RBF activation maps provides a full-brain
image, in MNI152 space, of the given principal component (Fig.~\componentBrains).

Next, we considered 80 topics estimated using Latent Dirichlet
Allocation~\citep{BleiEtal03} applied to 9,204 functional neuroimaging articles
in the Neurosynth database~\citep{RubiEtal17}. The topics, as well as
associated brain maps identified using Neurosynth, were identified and reported
in several prior studies~\citep{FoxEtal14, SulEtal17, ChenEtal20}. The topic
labels for each topic were generated automatically with the following ChatGPT
(chat.openai.com) prompt: ``Please help me come up with intuitive labels for
topic topics I found by fitting a topic model to thousands of neuroscience and
psychology articles. I'll paste in the top 10 highest-weighted words for each
topic, and I'd like you to respond with a suggested label. For each topic,
please respond with just the topic label and no other formatting or text. Here
are the next topic's top words:'' followed by a comma-separated list of the
given topic's top-weighted words reflected in the Table~\topics. For some
topics, ChatGPT responded with a longer-form response rather than a concise
topic label. In these instances, on a case-by-case basis, we used a second
follow-up prompt to achieve the given topic's label: ``Could you please come up
with a more concise label for that topic?''. We then manually identified a set
of 11 cognitive labels that were intended to encapsulate a representative range
of widely studied low-level and high-level cognitive functions. In choosing the
set of cognitive labels, we jointly considered each topic's ChatGPT-derived
topic label, along with the top-weighted words for the topic. We attempted to
generate a concise set of labels that still spanned the full set of cognitive
functions reflected across the 80 topics. Topics that appeared unrelated to
specific cognitive functions (e.g., topics related to specific methods or
clinical themes) are designated with dashes in Table~\topics.

@@

We manually identified 11 cognitive labels
spanning the set of 80 Neurosynth-derived topics: cognitive control, language
processing, memory, emotion, social cognition, spatial cognition, attention,
reward, sensory perception, motor control, and resting state. We then used
ChatGPT to automatically ``rank'' the processes from high-level to low-level
using the following prompt: ``Please rank these cognitive processes from
highest-level to lowest-level, where higher values indicate higher-order or
higher-level processes. Return the result as a csv file with a header row and
two columns: `Cognitive label' and `Rank'. Here are the processes: cognitive
control, language processing, memory, emotion, social cognition, spatial
cognition, attention, reward, sensory perception, motor control, resting
state''.  Table~\topicTags~displays the output.  We used these labels in
the analysis presented in Figure~\ref{fig:neurosynth-summary}E to help summarize
difference in topic weightings across experimental conditions.
